Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2021-09-22 analysis/draft_archive/paper_draft.Rmd
file cleaning

---
title: "Detecting Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
bibliography: references.bib
author:
  - name: "Sachin S. Pandya"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0001-7387-1307
  - name: "Xiaomeng Li"
    affiliation: "University of Connecticut | Ernest & Young"
  - name: "Timothy E. Moore"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0002-9576-0517
output: distill::distill_article
---

@@

With a peremptory challenge, a party declares a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why. While some jurisdictions have abolished peremptory strikes (Criminal Justice Act 1988, s 118(1) (UK); S.C. 2019, c. 25, s 269 (Canada)), federal and State law in the United States (except in Arizona^[Arizona Rule of Criminal Procedure XX]) authorize their use during jury selection in criminal and civil cases. In these cases, peremptory-challenge procedure can vary not only by State, but also by court and judge, including the number of peremptory challenges assigned to each side, the order of strikes, and the number of peremptory strikes each party may exercise at a time [@nscs-jury2018; @Williams2017].

@@

For a _Batson_ challenge, the challenging party must typically establish a "prima facie case", that is, enough evidence to infer that a party struck a juror because of such juror's race, gender, or other protected status. If so, the burden shifts to the challenged party to produce one or more nondiscriminatory reasons for that strike. Thereafter, the challenger must persuade the trial judge that prospective jurors' race or gender, not the proffered nondiscriminatory reason(s), was more likely than not the real motivation for the challenged strike(s) [@LaFave2020, sec. 22.3]. For similar challenges under State law, some States have reduced or eliminated the challenger's burden to establish a prima facie case [e.g., State v. Holloway, 209 Conn. 636, 646 & n.4 (1989); State v. Chapman, 317 S.C. 302, 305-06 (1995)], while other States have shifted the inquiry away from the striking party's motives to whether a hypothetical "objective observer" would find that race or ethnicity was a "factor" in that party's use of strikes [e.g., Wash. R. 37(e)].

@@

By law, evidence supporting a _Batson_ challenge may include not only the circumstances of the jury selection itself, but also data on peremptory challenges in past cases. For example, in a _Batson_ challenge against a prosecutor in a criminal case, such historical data would be relevant to whether (1) to conclude that the defendant has a prima facie case; (2) to believe a prosecutor from that office who has offered a race-neutral explanation for striking, say, an African-American juror in that case; and (3) to ultimately accept or reject a Batson challenge in that case [e.g., Miller-El v. Dretke, 545 U.S. 231, 253 (2005); Riley v. Taylor, 277 F.3d 261, 280-81, 283 (3d Cir. 2001); Cochran v. Herring, 43 F.3d 1404, 1412 (11th Cir. 1995)]. The US Supreme Court recently identified evidence for a _Batson_ challenge as including the "relevant history of the State's peremptory strikes in past cases" [Flowers v. Mississippi, 139 S. Ct. 2228, 2243 (2019)].

Prior work on statistical methods for _Batson_ challenges, however, has largely focused on strike data from a single case [e.g. @Barrett2007; @Gastwirth2016; @Kadane2018; @Kadane2018a], perhaps because strike data from past cases is hard to collect. Jury-selection records are often not readily available, because they are kept, if at all, by court staff somewhere other than the case file.  Such records are also largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise requires painstaking effort to collect [e.g., @Craft2018].

@@

When collecting strike data in multiple trials, prior studies have aggregated strikes across those trials, reported the observed difference in strike rates by race or gender, and tested for the probability of observing a non-zero difference in strike rates by race or gender, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference. For example, Eisenberg [-@Eisenberg2017, 334-339] used a 2-tail Fisher's exact test on the observed difference in peremptory strikes by race and gender in a non-random sample of thirty-five trials from 1997-2012 that resulted in death sentences (out of 63 such trials). @Gastwirth2013 has criticized the use of a paired t-test to compare the differences in strike rates of Black and non-Black venire members in the aggregate. Rather, both they and @Gastwirth2014 argued for applying the Cochran–Mantel–Haenszel test by taking the difference of the sum of the observed numbers of Black venire members challenged in each trial and the corresponding sum of the expected numbers of Blacks challenged, and then comparing it to the standard deviation of the difference.

Prior studies have also modeled how much a prospective juror's race affected the odds of being struck [e.g., @Baldus2001]. For exmpale, using jury selection data for 1992-2017 for Mississippi's Fifth Circuit Court District, Craft [-@Craft2018a, 9] applied a logistic regression model to analyze how much a prospective juror's race affected the odds of being struck by the state. For studies applying other modeling techniques to this Mississippi data, see @DeCamp2020 (propensity score matching) and @Dunn2021 (optimal feature selection, optimal classification trees).

@@

Second, we take that law as requiring, absent any data, the assumption that higher degrees of illegal bias (preference) are less likely than lower degrees of bias (preference). By using a normal (Gausian) distribution with $\sigma = 2$, we encode this assumption by making values of the bias parameter less likely as they move further from zero and by making very unlikely values any value of $b$ greater than two standard deviations from the mean.

@@

The posterior distribution does not have a closed form.  Accordingly, we used the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To measure convergence, we used traceplots, the Gelman-Rubin convergence diagnostic [@Gelman1992; @Gelman1998] and Geweke’s diagnostics [@Gewe1992].

@@

# INSERT simulation and plot code

load_path <- here::here("sim_resultc.csv")

sims$sig <- ifelse(sims$L80 <0 & sims$U80 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L80 >0 & sims$U80 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L80 <0 & sims$U80 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)

p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_minimal()+
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        #axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p

@@

```{r b95,  echo=FALSE,fig.cap="95% credible interval"}

@@

# Model 2: Simultaneous Strikes

@@

ggplot(data = b.com, aes(x = sim.n, y = b_s, color = sig))+
  facet_grid(~b)+
  geom_point(size = 3)+
  geom_errorbar(aes(x = sim.n, ymin = `2.5%`, 
                    ymax = `97.5%`), width = 0.4)+
  geom_hline(data = b.com, aes(yintercept = b.num), col = "red", lwd = 1.1) +
  geom_hline(yintercept = 0, col = "black", lwd = 1, lty = 1) +
  geom_vline(xintercept = 51.25, col = "black", lwd = 1, lty = 2) +
  xlab("Simulation")+
  scale_color_manual("Bias", values = c("blue", "grey"))+
  ylab("Posterior Bias Value")+
  ylim(c(-0.5, 5))+
  scale_x_continuous(breaks = c(1:50, 52.5), labels = c(1:50, "Average"))+
  coord_flip()+
  theme_classic()+
  theme(axis.title = element_text(size = 16), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9))
```
The points represent the estimated posterior mean and the lines represent the 95\% credible interval. The blue line means the credible interval does not include zero, thus indicating a bias, while the grey line means the credible interval includes zero and indicating no bias. We can find the estimated bias parameter of the 50 trials with different orders are close to each other for all of the three scenarios, indicating the order does not influence the estimate of the bias parameter. This is consistent with our previous premise and confirms that the simultaneous-strikes process can be modeled as a special case of our model of an alternating-strikes process.

@@

Following this suggestion, we describe here a prototype software application ("app") that implements the approach described above using `r R.version$version.string`[@R] and shiny [@shiny].

There is currently historical information of some prosecutors and defendants. More historical information would be updated when available. Users can input the current trial information as shown in table \@ref(tab:inpts), and then select the name of the prosecutor and the defendant from the pull-down window if applicable. If the name of the prosecutor or the defense attorney cannot be found, it means the historical information is not available.

The main screen displays two graphs -- one for the prosecution and the defense - that displays the prior density plot and the posterior density plot for the bias parameter.

```{r inpts, echo=FALSE}
df <- data.frame(round = c(1:10),
                 num_cog = c(4,4,4,3,2,2,2,2,2,2),
                 total = c(15,14,13,12,11,10,9,8,7,6),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))
knitr::kable(df)
```

For instance, if we collect the data of a new trial with prosecutor being "Brian P. Leaming" and defendant being "Frank J. Riccio II", and we would like to see if they are biased or not. We can select the name from the pull-down window of prosecutor and defendant to extract the historical information. Then we can input the current trial information, and select the weight that we would like to put on the historical information. Currently we provide three options for the weight, equal weight of historical information and current information, half weight of historical information and low weight of historical information. We can also select the cognizable class that we are interested in. There are currently two options of cognizable class, race and gender. Suppose the current trial information is as we show in table \@ref(tab:inpts), we put equal weight on historical information and we select cognizable class as race, then the posterior density as shown in figure \@ref(fig:brian) will be displayed.

@@

Sachin S. Pandya: Conceptualization, Investigation, Software, Writing; Xiaomeng Li: Conceptualization, Formal Analysis, Software, Writing - original draft; Timothy Moore: Conceptualization, Supervision, Visualization. Based on categories in https://casrai.org/credit/].

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - Sachin S. Pandya, University of Connecticut
#    orcid_id: 0000-0001-7387-1307
  - Xiaomeng Li, University of Connecticut | Ernest & Young
  - Timothy E. Moore, University of Connecticut
#    orcid_id: 0000-0002-9576-0517
#output: distill::distill_article
output: pdf_document
bibliography: references.bib
abstract: |
  US constitutional law bars using peremptory strikes during jury selection because of race, gender, or another cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias based on an attorney's pattern of peremptory challenges in both a current trial as well that attorney's use of strikes in past cases. Specifically, we show how to use historical data on an attorney's use of peremptory strikes in past trials to estimate, in a current trial, how likely an attorney has used peremptory challenges because of the cognizable class of prospective jurors. We also introduce the use of a _power prior_ to provide flexibility in the use of such historical data. We show the utility of our model on simulated data. Finally, we extend our approach by presenting an R Shiny application that can be used to examine jury selection data in real time. We demonstrate the app using actual jury-selection data from a convenience sample of criminal trials in the federal district court of Connecticut from YYYY to YYYY.

---

```{r setup, include=FALSE}

library(ggplot2)
library(dplyr)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

@@

With a peremptory challenge, a party declares a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.^[While some jurisdictions have abolished peremptory strikes (Criminal Justice Act 1988, s 118(1) (UK); S.C. 2019, c. 25, s 269 (Canada)), federal and State law in the United States authorize their use during jury selection in criminal and civil cases. The current exception is Arizona. Arizona Rule of Criminal Procedure XX.] Peremptory-challenge procedure can vary not only by State, but also by court and judge, including the number of peremptory challenges assigned to each side, the order of strikes, and the number of peremptory strikes each party may exercise at a time [@nscs-jury2018; @Williams2017].

This paper focuses on _Batson_ challenges to a party's use of peremptory challenges.

@@

For a _Batson_ challenge, the challenging party must typically establish a "prima facie case", that is, enough evidence to infer that a party struck a juror because of such juror's race, gender, or other protected status. If so, the burden shifts to the challenged party to produce one or more nondiscriminatory reasons for that strike. Thereafter, the challenger must persuade the trial judge that prospective jurors' race or gender, not the proffered nondiscriminatory reason(s), was more likely than not the real motivation for the challenged strike(s) [@LaFave2020, sec. 22.3].

For some similar challenges under State law, some States have reduced or eliminated the challenger's burden to establish a prima facie case [e.g., State v. Holloway, 209 Conn. 636, 646 & n.4 (1989); State v. Chapman, 317 S.C. 302, 305-06 (1995)], while other States require proving not motive, only that an "objective observer" would find that race or ethnicity was a "factor" in that party's use of strikes [e.g., Wash. R. 37(e)].

@@

By law, evidence supporting such challenges may include aggregate data on peremptory challenges in past cases. For example, in a _Batson_ challenge against a prosecutor in a criminal case, such historical data would be relevant to whether (1) to conclude that the defendant has a prima facie case; (2) to believe a prosecutor from that office who has offered a race-neutral explanation for striking, say, an African-American juror in that case; and (3) to ultimately accept or reject a Batson challenge in that case [e.g., Miller-El v. Dretke, 545 U.S. 231, 253 (2005); Riley v. Taylor, 277 F.3d 261, 280-81, 283 (3d Cir. 2001); Cochran v. Herring, 43 F.3d 1404, 1412 (11th Cir. 1995)]. The US Supreme Court recently identified evidence for a _Batson_ challenge as including the "relevant history of the State's peremptory strikes in past cases" [Flowers v. Mississippi, 139 S. Ct. 2228, 2243 (2019)].

Prior work on statistical methods for _Batson_ challenges, however, has largely focused on strike data from a single case [e.g. @Barrett2007; @Gastwirth2016; @Kadane2018; @Kadane2018a], perhaps because strike data from past cases is hard to collect. Jury-selection records are often not readily available, because they are kept, if at all, by court staff somewhere other than the case file.  Such records are also largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise requires painstaking effort to collect [e.g., @APM2018].

@@

When collecting strike data in multiple trials, prior studies have aggregated strikes across those trials, reported the observed difference in strike rates by race or gender, and tested for the probability of observing a non-zero difference in strike rates by race or gender, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference.

For example, Eisenberg [-@Eisenberg2017, 334-339] used a 2-tail Fisher's exact test on the observed difference in peremptory strikes by race and gender in a non-random sample of thirty-five trials from 1997-2012 that resulted in death sentences (out of 63 such trials). @Gastwirth2013 has criticized the use of a paired t-test to compare the differences in strike rates of Black and non-Black venire members in the aggregate. Rather, both they and @Gastwirth2014 argued for applying the Cochran–Mantel–Haenszel test by taking the difference of the sum of the observed numbers of Black venire members challenged in each trial and the corresponding sum of the expected numbers of Blacks challenged, and then comparing it to the standard deviation of the difference.

Prior studies have also modeled how much a prospective juror's race affected the odds of being struck [e.g., @Baldus2001]. For example, using jury selection data for 1992-2017 for Mississippi's Fifth Circuit Court District, Craft [-@Craft2018a, 9] applied a logistic regression model to analyze how much a prospective juror's race affected the odds of being struck by the state. For studies applying other modeling techniques to this Mississippi data, see @DeCamp2020 (propensity score matching) and @Dunn2021 (optimal feature selection, optimal classification trees).

@@

Second, we take that law as requiring, absent any data, the assumption that higher degrees of illegal bias (preference) are less likely than lower degrees of bias (preference). By using a normal (Gaussian) distribution with $\sigma = 2$, we encode this assumption by making values of the bias parameter less likely as they move further from zero and by making very unlikely values any value of $b$ greater than two standard deviations from the mean.

@@

The posterior distribution does not have a closed form.  Accordingly, we used the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To measure convergence, we used traceplots, the Gelman-Rubin convergence diagnostic [@Gelman1992; @Gelman1998] and Geweke’s diagnostics [@Geweke1992].

@@

p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal() +
  theme_minimal() +
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        #axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p

@@

```{r b95, fig.cap="95% credible interval"}

@@

# Extension: Simultaneous Strikes

@@

ggplot(data = b.com, aes(x = sim.n, y = b_s, color = sig))+
  facet_grid(~b)+
  geom_point(size = 3)+
  geom_errorbar(aes(x = sim.n, ymin = `2.5%`, 
                    ymax = `97.5%`), width = 0.4)+
  geom_hline(data = b.com, aes(yintercept = b.num), col = "red", lwd = 1.1) +
  geom_hline(yintercept = 0, col = "black", lwd = 1, lty = 1) +
  geom_vline(xintercept = 51.25, col = "black", lwd = 1, lty = 2) +
  xlab("Simulation")+
  scale_color_manual("Bias", values = c("blue", "grey"))+
  ylab("Posterior Bias Value")+
  ylim(c(-0.5, 5))+
  scale_x_continuous(breaks = c(1:50, 52.5), labels = c(1:50, "Average"))+
  coord_flip()+
  theme_classic()+
  theme(axis.title = element_text(size = 16), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9))

The points represent the estimated posterior mean and the lines represent the 95\% credible interval. The blue line means the credible interval does not include zero, thus indicating a bias, while the grey line means the credible interval includes zero and indicating no bias. We can find the estimated bias parameter of the 50 trials with different orders are close to each other for all of the three scenarios, indicating the order does not influence the estimate of the bias parameter. This is consistent with our previous premise and confirms that the simultaneous-strikes process can be modeled as a special case of our model of an alternating-strikes process.

@@

Following this suggestion, we describe here a prototype software application ("app") that implements the approach described above using `r R.version$version.string`[@R] and shiny [@shiny] and comes loaded with real jury-selection data from a convenience sample of criminal cases in the federal district court for Connecticut during YYYY through YYYY. To partially de-identify this historical data, we excluded defendants' names and substituted fictitious names for the attorneys using @charlatan.

With the app, users can input the current information on the use of peremptory strikes as shown in table \@ref(tab:inpts), and then select the name of the prosecutor and the defendant from the pull-down window, if available, in the historical data. If an attorney's name cannot be found, it means that the app does not have any historical information on that attorney's use of peremptory strikes.

The main screen displays two graphs -- one for the prosecution and the defense - that displays the prior and posterior density plots for the bias parameter.

```{r inpts, echo=FALSE, eval = FALSE}
df <- data.frame(round = c(1:10),
                 num_cog = c(4,4,4,3,2,2,2,2,2,2),
                 total = c(15,14,13,12,11,10,9,8,7,6),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))
knitr::kable(df)

For instance, if we collect the data of a new trial with prosecutor being "Brian P. Leaming" and defendant being "Frank J. Riccio II", and we would like to see if they are biased or not. We can select the name from the pull-down window of prosecutor and defendant to extract the historical information. Then we can input the current trial information, and select the weight that we would like to put on the historical information. Currently we provide three options for the weight, equal weight of historical information and current information, half weight of historical information and low weight of historical information. We can also select the cognizable class that we are interested in. There are currently two options of cognizable class, race and gender.

Suppose the current trial information is as we show in table \@ref(tab:inpts), we put equal weight on historical information and we select cognizable class as race, then the posterior density as shown in figure \@ref(fig:brian) will be displayed.

@@

Sachin S. Pandya: Conceptualization, Investigation, Software, Writing; Xiaomeng Li: Conceptualization, Formal Analysis, Software, Writing - original draft; Timothy Moore: Conceptualization, Supervision, Visualization, Writing. Based on categories in https://casrai.org/credit/].
